The role of aneuploidy in the evolution of cancer drug resistance

### 2024-02-01 EvolutionaryRescue1.tex
Add files via upload

\paragraph{Aneuploidy in cancer.}  Each year approximately 10 million people die from cancer \citep{kocarnik2022cancer}, so understanding that factors that contribute to failure of interventions is of great importance. One hypothesized factor is aneuploidy, where cells are characterized by an imbalanced karyotype and chromosome copy number alterations \citep{schukken2018cin}, caused by chromosomal instability (CIN), the mitotic process in which cells suffer from chromosome mis-segregation that leads to aneuploidy. Importantly, aberrations in chromosome copy number have been shown to allow cancer cells to survive under stressful conditions such as drug therapy \citep{lukow2021chromosomal,rutledge2016selective}. Indeed, cancer cells are often likely to be aneuploid, and aneuploidy is associated with poor patient outcomes \citep{ben2020context,smith2018systematic}.

@@

\paragraph{Bootstrapping}
We  use bootstrap resampling method in order to compute  confidence intervals for the mean threshold tumor sizes and mean times by resampling 100 times with replacement from the data set obtain from simulations. We then compute the interval which contain $95\%$ of the means of each resampled set.

@@

\paragraph{Directions of future research}
Even though our model has been designed for cancer it can be extended to understand to understand evolutionary in different biological contexts, for example, how yeast subject to stress can overcome extinction with the help of aneuploidy~\citep{pompei2023fitness}. Additionally, the heterogeneity of aneuploidy is a fact the we did not account for, as not all the aneuploidy lineages generated have the same growth rate $\Delta_a$. This can be corrected by  fixing the aneuploidy death rates and sampling the aneuploidy growth rates from a distribution $f\left(\Delta_a|\mu_a\right)$~\citet{martin2013probability}. Furthermore, tumor heterogeneity is also a factor for pre-existing genetic variation where the fitness of the aneuploid cells can be drawn from a distribution upon which selection acts when the tumor is exposed to a strong stress.

@@

\caption{The probability of evolutionary rescue (i.e. the probability that the population does not go to extinction), $\presc$, as a function of the initial tumor size, $N$. Dashed vertical line shows the threshold tumor size, above which the probability is very high. Blue dashed line represents the probability of evolutionary rescue as a function of $N$ without aneuploidy ($u=0$). The black line represents the case with tolerant aneuploidy a fraction $f=0\%$ of the initial tumor being aneuploid, the red line represents the case with a fraction $f=5\%$ and the green line represents the case with $f=50\%$. The dots represent simulations and the error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the probability of that the tumor has adapted to the stress and $n=100$ is the number of simulations. Parameters: $\lambda_w=0.1,\lambda_a=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$.}

@@

Our inhomogeneous Poisson-process approximation (solid black line, \cref{meanproliferationtime}) is in agreement with simulation results (red markers with 95\% confidence intervals obtained with bootstrapping, see Appendix G) for small and intermediate values of $N$. Dashed vertical blue line represents the threshold tumor size above which evolutionary rescue is very likely through aneuploidy \cref{eq:N_a} and the dashed vertical green line represents the threshold tumor size above which evolutionary rescue is very likely through direct mutation \cref{eq:N_m}. Solid lines represents the approximations \cref{eq:AsymptoticTimeRulesRecurrence} ($N<N_a^*$ blue line and $N>N_m^*$ green line). The simulations converge to the solution of \cref{eq:t2det} (green line) for large values of $N\gg N_m^*$.  Parameters: $\lambda_w=0.1,\lambda_a=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$.}

@@

Our inhomogeneous Poisson-process approximation (solid black line, \cref{meanproliferationtime}) is in agreement with simulation results (red markers with 95\% confidence intervals obtained with bootstrapping, see Appendix G) for small and intermediate values of $N$. Dashed vertical blue line represents the threshold tumor size above which evolutionary rescue is very likely through aneuploidy \cref{eq:N_a} and the dashed vertical green line represents the threshold tumor size above which evolutionary rescue is very likely through direct mutation \cref{eq:N_m}. Solid lines represents the approximations \cref{eq:AsymptoticTimeRulesRecurrence} ($N<N_a^*$ blue line and $N>N_m^*$ green line). The simulations converge to the solution of \cref{eq:t2det} (green line) for large values of $N\gg N_m^*$.  Parameters: $\lambda_w=0.1,\lambda_a=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$.}

@@

\caption{Shown is the mean time for appearance of a resistance mutation the leads to evolutionary rescue with aneuploidy ($u>0$). Our inhomogeneous Poisson-process approximations (solid black lines, right: \cref{meantimet2}). Dashed black line is the  inhomogeneous Poisson-process approximation \cref{meantimet2SGV} where a fraction $f$ of tumor is aneuploid at the onset of drug therapy which is in agreement with simulation results (red markers with 95\% confidence intervals obtained with bootstrapping, see Appendix G). Dashed vertical blue line represents the threshold tumor size above which evolutionary rescue is very likely through aneuploidy \cref{eq:N_a} and the dashed vertical green line represents the threshold tumor size above which evolutionary rescue is very likely through direct mutation \cref{eq:N_m}. Solid lines represents the approximations \cref{eq:AsymptoticTimeRules} ($N<N_a^*$ blue line and $N>N_m^*$ green line). 

>>>>

\paragraph{Aneuploidy in cancer.}  Each year approximately 10 million people die from cancer \citep{kocarnik2022cancer}, so understanding the factors that contribute to failure of interventions is of great importance. One hypothesized factor is aneuploidy, where cells are characterized by an imbalanced karyotype and chromosome copy number alterations \citep{schukken2018cin}, caused by chromosomal instability (CIN), the mitotic process in which cells suffer from chromosome mis-segregation that leads to aneuploidy. Importantly, aberrations in chromosome copy number have been shown to allow cancer cells to survive under stressful conditions such as drug therapy \citep{lukow2021chromosomal,rutledge2016selective}. Indeed, cancer cells are often likely to be aneuploid, and aneuploidy is associated with poor patient outcomes \citep{ben2020context,smith2018systematic}.

@@

\paragraph{Bootstrapping}
We  use bootstrap resampling method in order to compute  confidence intervals for the mean threshold tumor sizes and mean times by resampling 100 times with replacement from the data set obtained from simulations. We then compute the interval which contain $95\%$ of the means of each resampled set.

@@

\paragraph{Directions of future research}
Even though our model has been designed for cancer it can be extended to understand evolutionary rescue in different biological contexts, for example, how yeast subject to stress can overcome extinction with the help of aneuploidy~\citep{pompei2023fitness}. Additionally, the heterogeneity of aneuploidy is a fact the we did not account for, as not all the aneuploidy lineages generated have the same growth rate $\Delta_a$. This can be corrected by  fixing the aneuploidy death rates and sampling the aneuploidy growth rates from a distribution $f\left(\Delta_a|\mu_a\right)$~\citet{martin2013probability}. Furthermore, tumor heterogeneity is also a factor for pre-existing genetic variation where the fitness of the aneuploid cells can be drawn from a distribution upon which selection acts when the tumor is exposed to a strong stress.

@@

\caption{The probability of evolutionary rescue (i.e. the probability that the population does not go to extinction), $\presc$, as a function of the initial tumor size, $N$. Dashed vertical line shows the threshold tumor size, above which the probability is very high. Blue dashed line represents the probability of evolutionary rescue as a function of $N$ without aneuploidy ($u=0$). The black line represents the case where a fraction $f=0\%$ of the initial tumor is aneuploid, the red line represents the case with $f=5\%$ and the green line represents the case with $f=50\%$. The dots represent simulations and the error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the fraction of simulations in which the tumor has adapted to the stress and $n=100$ is the number of simulations. Parameters: $\lambda_w=0.1,\lambda_a=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$.}

@@

Our inhomogeneous Poisson-process approximation (solid black line, \cref{meanproliferationtime}) is in agreement with simulation results (red markers with 95\% confidence intervals obtained with bootstrapping, see Appendix G) for small and intermediate values of $N$. Dashed vertical blue line represents the threshold tumor size above which evolutionary rescue is very likely through aneuploidy \cref{eq:N_a} and the dashed vertical green line represents the threshold tumor size above which evolutionary rescue is very likely through direct mutation \cref{eq:N_m}. Solid lines represents the approximations \cref{eq:AsymptoticTimeRulesRecurrence} ($N<N_a^*$ blue line and $N>N_m^*$ green line). The simulations converge to \cref{eq:t2det} (green line) for large values of $N\gg N_m^*$.  Parameters: $\lambda_w=0.1,\lambda_a=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$.}

@@

\caption{The mean time for the mutant cell population to reach size $M$, where $M$ is the tumor detection size.
Our inhomogeneous Poisson-process approximation (solid black line, \cref{meanproliferationtime2}) is in agreement with simulation results (red markers with 95\% confidence intervals obtained with bootstrapping, see Appendix G) for small and intermediate values of $N$. Dashed vertical blue line represents the threshold tumor size above which evolutionary rescue is very likely through aneuploidy \cref{eq:N_a} and the dashed vertical green line represents the threshold tumor size above which evolutionary rescue is very likely through direct mutation \cref{eq:N_m}. Solid blue line represents the approximation \cref{meanproliferationtime2} with $\tau_a$ from \cref{eq:AsymptoticTimeRules} for $N<N_a^*$ and the solid green line  represents the approximation \cref{eq:t2det2} for $N>N_m^*$. The simulations converge to \cref{eq:t2det2} (green line) for large values of $N\gg N_m^*$.  Parameters: $\lambda_w=0.1,\lambda_a=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$.}

@@

\caption{Shown is the mean time for appearance of a resistance mutation the leads to evolutionary rescue with aneuploidy ($u>0$). Black lines represent our inhomogeneous Poisson-process approximations (solid black line, \cref{meantimet2}; dashed black line \cref{meantimet2SGV}). Dashed black line is the  inhomogeneous Poisson-process approximation where a fraction $f$ of tumor is aneuploid at the onset of drug therapy which is in agreement with simulation results (red markers with 95\% confidence intervals obtained with bootstrapping, see Appendix G). Dashed vertical blue line represents the threshold tumor size above which evolutionary rescue is very likely through aneuploidy \cref{eq:N_a} and the dashed vertical green line represents the threshold tumor size above which evolutionary rescue is very likely through direct mutation \cref{eq:N_m}. Solid lines represents the approximations \cref{eq:AsymptoticTimeRules} ($N<N_a^*$ blue line and $N>N_m^*$ green line). 
